---
title: "The Mini Android"
author: "Inés Añíbarro Gorostiza"
date: "October 2018"
output:
  pdf_document: default
params:
  test: 'default'
---

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
#Loading the packages
#ipak <- function(pkg){
#  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
#  if (length(new.pkg))
#    install.packages(new.pkg, dependencies = TRUE)
#  sapply(pkg, require, character.only = TRUE)
#}

#packages <- c('fBasics', 'fUnitRoots', 'forecast', 'nnet', 'zoo')
#ipak(packages)

library(fBasics)
library(forecast)
library(nnet)
library(zoo)
```

```{r, echo=FALSE}
# test if there is at least one argument: if not, return an error
if (length(args)==0) {
  stop("At least one argument must be supplied (input file).\n", call.=FALSE)
} 

#data <-read.csv(params$args, header = TRUE)
data <- params$test
```

```{r, echo = FALSE}
#first thing to check: is the data is constant ? if so, we cannot do much!
constant <- (length(unique(na.omit(data[,2]))) == 1)
#period set to 1, to avoid duplication of eval arguments in the chunks
if (constant){
  period <- 1
  pValue <- 0
  typ <- 'null'
}
```


```{r, echo=FALSE, eval = (constant == FALSE)}
#cuidado: lo primero que tengo que mirar es la longitud del periodo para poder crear la serie 
acf <- as.vector(acf(data[,2], plot = FALSE, na.action = na.pass )$acf)
```

1) Verificacion: el forecast package va bien!!

```{r, echo = FALSE, eval = (constant == FALSE)}
period <- findfrequency(data[,2])
```

`r if (period != 1) {paste('This method returns a period of length ', period, '. We have found, empirically, that in practice, this method performs better than the previous one. We will keep this result for the future.', sep = '')}`

```{r, echo=FALSE}
ts <- ts(data[,2], frequency = period)
name <- colnames(data)[2]
```


#Time series to evaluate: `r name`

### 1) Esto es una pruebilla para ver que opencpu funciona!!! sorpresa!!!

This time series represents the evolution of the number of `r name`. 

```{r, echo = FALSE}
plot(ts, ylab = name)
```


It contains `r length(ts)` observations, going from `r as.POSIXct(data[1,1]/1000, origin="1970-01-01")` to `r as.POSIXct(data[length(data[,1]),1]/1000, origin="1970-01-01")`. 

```{r, echo = FALSE, eval = (constant == TRUE)}
const <- unique(na.omit(data[,2]))
```

`r if(constant == TRUE){paste('The values of the time series are constant, all of them are equal to ', unique(na.omit(data[,2])), '. Unfortunately, a constant series cannot be analyzed much more and so we end up the report here.', separator = '')}`


2) Verificacion: la libreria zoo funciona

```{r, echo = FALSE, eval = (constant == FALSE)}
nas <- sum(is.na(ts))

if(nas == 0){
  chunki <- 'This time series does **not** contain any **missing values**, thus there is no need to make any interpolation.'
}

if (nas != 0){
  chunki <- paste('This time series contains ', nas, ' **missing values**. We will interpolate NAs linearly with the values that wrap the missing value. Missing values in the extremes of the series (first or last values) will be refilled with the closest point.')
  
  ts <- na.approx(ts, rule = 2)
}

min_ts <- min(ts)
max_ts <- max(ts)
mean_ts <- mean(ts)
```

`r if (constant == FALSE){paste('The time series contains values between ', min_ts, ' and ', max_ts, ' with a mean of ', round(mean_ts,3), '.', separator = '')}`

`r if (constant == FALSE){chunki}`

We will normalize the series in order to have all values between 0 and 1. This will be useful when comparing it with other time series that have different units. We display the plot of the normalized series.

```{r, echo = FALSE}
ts <- (ts - min_ts)/(max_ts - min_ts)
plot(ts, ylab = name)
```

